---
subtitle: "MA8701"
title: "Project 1"
author: "Group 5 : Yellow Submarine"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  #html_document
  pdf_document
---
  
```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE,tidy=TRUE,message=FALSE,warning=FALSE,strip.white=TRUE,prompt=FALSE,
                      cache=TRUE, size="scriptsize",fig.width=4, fig.height=3)
```

```{r rpackages,eval=TRUE,echo=FALSE}
#load packages
library(GGally)
library(tidyverse)
library(glmnet)
library(grplasso)
library(corrplot)
library(RColorBrewer)
library(leaps)
```

# The Data Set 

For our project work we use the Munich Rent 2003 data set as described in https://rdrr.io/cran/LinRegInteractive/man/munichrent03.html.

The data set has the covariates
-   `nmqm`: rent per square meter (numeric)
-   `wfl`: area in square meters (numeric)
-   `rooms`: number of rooms (numeric)
-   `bj`: year of construction (factor)
-   `bez`: district (factor)
-   `wohngut`: quality of location (factor)
-   `wohnbest`: high quality of location (factor)
-   `ww0`: hot water supply available (factor)
-   `zh0`: central heating (factor)
-   `badkach0`: tiled bathroom (factor)
-   `badextra`: high-quality bathroom (factor)
-   `kueche`: upscale kitchen equipment (factor)
and the response
-   `nm`: rental price (numeric).

```{r loadData,eval=TRUE,echo=FALSE}
# Load data set
munich_house <- read.table(
  "https://data.ub.uni-muenchen.de/2/1/miete03.asc",
  sep="\t", header=TRUE)
M <- cor(munich_house)
corrplot(M, type = "upper")
# corrplot(M, type = "upper", order = "hclust", col = brewer.pal(n = 8, name = "RdYlBu"))
# Transform bez and bj to factor since interpreted as numeric by read.table
munich_house$bez=as.factor(munich_house$bez) 
munich_house$bj=as.factor(munich_house$bj)
str(munich_house)
```

We store the data set in an R data frame for all further computations. 

```{r buildDataFrame,eval=TRUE,echo=TRUE} 
# Create model matrix without intercept and nmqm 
# (since nmqm*wfl=nm gives the response directly)
x_mod <- model.matrix(nm~.-nmqm,data=munich_house)[,-1] 
y_mod <- munich_house$nm
df_mod <- data.frame(y_mod,x_mod)
colnames(df_mod)[1]="nm"
```


# Regression

We start with a vanilla regression for reference.

```{r regression, eval=TRUE, echo=FALSE}
# Vanilla regression model
lm_mod <- lm(nm~.,data=df_mod) 
#Printing the summary to assess significant parameters
summary(lm_mod)
```

Remark: Interestingly in the regression, the significance of different `bj`s and `bez`s varies a lot.  

# Shrinkage

After we saw the results for the linear regression, we continue with shrinkage methods.

## Subset selection

```{r subset, eval = TRUE, echo = TRUE}
munich_house <- read.table(
  "https://data.ub.uni-muenchen.de/2/1/miete03.asc",
  sep="\t", header=TRUE)
x <- apply(munich_house[, -c(1, 2)], 2, scale) # standardize input
y <- munich_house[, 1] -  mean(munich_house[, 1]) # centralize output to remove intercept
data <- data.frame(y, x)
bests <- regsubsets(x, y)
sumbests <- summary(bests)
print(sumbests)
which.min(sumbests$cp)
lm_sub <- lm(y ~ wfl + bj + bez + wohngut + ww0 + badkach0 + kueche, data = data)
summary(lm_sub)
confint(lm_sub)

```

## Ridge

```{r ridge, eval = TRUE, echo = TRUE}
munich_house <- read.table(
  "https://data.ub.uni-muenchen.de/2/1/miete03.asc",
  sep="\t", header=TRUE)
x <- apply(munich_house[, -c(1, 2)], 2, scale) # standardize input
y <- munich_house[, 1] -  mean(munich_house[, 1]) # centralize output to remove intercept
data <- data.frame(y, x)
start <- glmnet(x = x, y = y, alpha = 0)
autolambda <- start$lambda
newlambda <- c(autolambda, 10, 5, 3, 1, 0.5, 0.1) # add more to approach zero lambda
ridge_fit <- glmnet(x, y, alpha = 0, lambda = newlambda)
plot(ridge_fit, xvar = "lambda", label = T)
cv.ridge <- cv.glmnet(x, y, alpha = 0, lambda = newlambda)
print(paste("The lamda giving the smallest CV error",cv.ridge$lambda.min))
print(paste("The 1sd err method lambda",cv.ridge$lambda.1se))
plot(cv.ridge)
plot(ridge_fit, xvar = "lambda", label = T) + abline(v = log(cv.ridge$lambda.1se))
coef(cv.ridge)

```


## Lasso

```{r lasso, eval=TRUE, echo=FALSE}
gen_mod <- glmnet(x=x_mod,y=y_mod,alpha=1,standardize=TRUE)
gen_lambda_mod <- gen_mod$lambda
plot(gen_lambda_mod) 

plot(gen_mod) #plot against the L1 norm
plot(gen_mod,xvar="lambda",label=TRUE)
# abline(v=log(cv_gen_mod$lambda.1se)) #use 1sd error rule default

# results_mod=cbind(coef(gen_mod,s=cv_gen_mod$lambda.1se),coef(lm_mod))
# colnames(results_mod)=c("general lasso","vanilla LS")
# print(results_mod)
```

For the $\lambda$ with one standard deviation, we observe that many of the `bj`s and `bez`s get shrinked, but not all of them - and the values differ from the linear regression. Whereas the other kept covariants roughly keep their parameter.

Above we considered a fixed $\lambda$, now we analyse which $\lambda$ is optimal using cross validation.

```{r lassoCV, eval=TRUE, echo=FALSE}
# Choosing optimal lambda performing Crossvalidation
cv_gen_mod=cv.glmnet(x=x_mod,y=y_mod,alpha=1)
print(paste("The lamda giving the smallest CV error",cv_gen_mod$lambda.min))
print(paste("The 1sd err method lambda",cv_gen_mod$lambda.1se))
plot(cv_gen_mod)
```


## Group lasso

```{r groupLasso, eval=TRUE, echo=FALSE}
# Adding an intercept to the design matrix 
# NB! In contrast to glm_net the grplasso package requires the intercept
x_mod_group <- cbind(1, x_mod)
colnames(x_mod_group)[1] <- c("Intercept")
# Building new data frame with intercept now for the group lasso
df_mod_group <- data.frame(y_mod,x_mod_group)
colnames(df_mod_group)[1] <-c("nm")

# Defining the grouped classes
district <- rep(4,24) #25 levels of factor bez, why 24? This must be changed according to the design matrix
year <- rep(3,43) #44 levels of factor bj, has to be 43?
index_mod <- c(NA,1,2,year,district,5,6,7,8,9,10,11) #individual groups for all except bez and bj

grp_lambda_mod <- lambdamax(x=x_mod_group, y = y_mod, index = index_mod, penscale = sqrt,
                          model = LinReg(),standardize = TRUE,center = TRUE)*0.5^(0:20)

# Fit using grouped index
# NB! Neglecting first 3 lambdas for better scale in the plot
grp_mod <- grplasso(x=x_mod_group, y = y_mod, index = index_mod, lambda = grp_lambda_mod[-(1:3)], model = LinReg(),
                    penscale = sqrt,
                    control = grpl.control(update.hess = "lambda", trace = 0))


## Plot coefficient paths
plot(grp_mod) 
str(grp_mod)
coef(grp_mod) 

```


In the grouped lasso, the `bj` and `bez` are all shrinked or are all inclueded, respectively. This coincides better with our intuition, that this criterion is considered or not considered. Whereas in the regression and lasso before, just some years of construction and some areas where significant.


.............. STILL TO DO BY FB!!!!!


```{r final, eval=TRUE, echo=FALSE}
#CV for å finne optimal lambda? Bootstrapping? 
library(gglasso)
#package chosen lambdagrid
fitls_mod <- gglasso(x = x_mod_group[,-1], y = y_mod, group = index_mod[-1], loss = "ls")
plot(fitls_mod)
#own defined lambdagrid
fitls_mod <- gglasso(x = x_mod_group[,-1], y = y_mod, group = index_mod[-1], loss = "ls",lambda=grp_lambda_mod)
plot(fitls_mod)

#crossvalidation on self defined lambdagrid - this takes some time
cvfitls_mod <- cv.gglasso(x = x_mod_group[,-1], y = y_mod, group = index_mod[-1], loss = "ls",lambda=grp_lambda_mod)
plot(cvfitls_mod)
str(cvfitls_mod)
cvfitls_mod$lambda.1se

#fit with optimal lambda on defined lambda grid(either package made or made by me)
fin_mod=gglasso(x = x_mod_group[,-1], y = y_mod, group = index_mod[-1], loss = "ls",lambda=cvfitls_mod$lambda.1se)
coef(fin_mod)

#testing with a fixed lambda
test_mod=gglasso(x = x_mod_group[,-1], y = y_mod, group = index_mod[-1], loss = "ls",lambda=5)
coef(test_mod)

#final results, group lasso either shrinks all or nothing and even increases some estimated parameter coefficients, what do we take away from this? 
results_mod=cbind(coef(fin_mod),coef(gen_mod,s=cv_gen_mod$lambda.1se),coef(lm_mod))
colnames(results_mod)=c("group lasso","general lasso","vanilla LS")
print(results_mod)

```